# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
```{r, echo=TRUE}
unzip("./activity.zip") # contains activity.csv
data <- read.csv("./activity.csv")
head(data)
```

## What is mean total number of steps taken per day?

First, let's see a histogram.
```{r, fig.width=8, fig.height=4, echo=TRUE}
steps.each.day <- tapply(data$steps, data$date, sum)

library(ggplot2)
bwidth <- diff(range(steps.each.day, na.rm=TRUE)) / 30
qplot(steps.each.day, geom="histogram", binwidth=bwidth) +
    geom_rug(color="blue", alpha=0.7) +
    labs(x="Steps per day", y="Frequency", title="Histogram of total steps per day")

mu <- mean(steps.each.day, na.rm=TRUE)
med <- median(steps.each.day, na.rm=TRUE)
mu
med
```

The mean number of total steps per day is **`r mu`** and the median
is **`r med`**.


## What is the average daily activity pattern?

The daily activity pattern averaged within each five minute interval across all
days:
```{r, fig.width=8, fig.height=4, echo=TRUE}
library(plyr)
interval.means <- ddply(data, "interval", summarise,
                        mean=mean(steps, na.rm=TRUE))
ggplot(interval.means, aes(interval, mean)) + geom_path() +
    labs(title="Mean # Steps by 5-minute Time Interval", x="Interval")

idx <- which.max(interval.means$mean)
interval.means[idx,]
```

The interval with the maximum number of steps when averaged across all days is
**interval `r interval.means[idx, "interval"]`**.

## Imputing missing values
```{r, echo=TRUE}
incomplete.rows <- sum(!complete.cases(data))
incomplete.rows
```
It looks like we have **`r incomplete.rows`** rows with NAs in our dataset.
Before we start with imputation, let's try to get a feel for where the missing
values are. We can start by counting how many NA values exist within each five
minute interval across all days using ddply:

```{r, echo=TRUE}
interval.nas <- ddply(data, "interval", summarise, na.count=sum(is.na(steps)))
head(interval.nas)
```

Now isn't that interesting? The first few values are 8. I wonder how many
intervals have 8 NAs, could it be all of them?

```{r, echo=TRUE}
all(interval.nas$na.count == 8)
```

Indeed! My guess is that this means that there were 8 days where no data were
collected. Let's confirm. We'll use ddply again to count the total
number of NA observations by day:

```{r, echo=TRUE}
intervals.per.day <- 24 * 60 / 5 # how many 5 minute intervals in a day (288)
intervals.per.day
date.nas <- ddply(data, "date", summarise, na.count=sum(is.na(steps)))
head(date.nas)
sum(date.nas$na.count == intervals.per.day)
```

As expected, there are eight days with no data, and furthermore these eight
days contain the entire set of NAs:

```{r, echo=TRUE}
dates.with.nas <- unique(date.nas$date[date.nas$na.count != 0])
length(dates.with.nas)
```

Given what we have learned so far, I believe that a reasonable imputation
strategy would be to populate the days that are missing data with the mean
values for that day of the week.

To get started, let's add a **weekday** column to our original data frame:

```{r, echo=TRUE}
data$weekday <- factor(weekdays(as.Date(data$date)),
                       levels=c("Sunday", "Monday", "Tuesday", "Wednesday",
                                "Thursday", "Friday", "Saturday"))
head(data)
head(data[data$interval == 0,])
weekdays.with.na <- unique(data[data$date %in% dates.with.nas,]$weekday)
weekdays.with.na
```

Looks good, now we need to compute the mean within each five minute interval
separately for each weekday:

```{r, echo=TRUE}
means.sd <- ddply(data, c("interval", "weekday"),
                  summarise, steps=mean(steps, na.rm=TRUE))
```

Now let's question whether or not computing the mean by weekday is going to be
any more informative than just substituting the overall mean across all days
(this is a bit of foreshadowing for the next question):

```{r, fig.width=8, fig.height=8, echo=TRUE}
ggplot(means.sd, aes(interval, steps)) + geom_line() +
    facet_grid(weekday ~ .) +
    labs(x="Interval", y="Steps", title="Daily activity")
```

There seems to be enough variability between days to support imputation by
weekday. Let's fill in the missing values.

```{r, echo=TRUE}
# Separate the complete and incomplete cases
complete.ind <- complete.cases(data)
complete <- data[complete.ind,]
incomplete <- data[!complete.ind,]
# Merge the incomplete cases with the per-interval means by weekday
merged <- merge(incomplete[, c("weekday", "interval", "date")], means.sd,
                by=c("interval", "weekday"))
# Bind the complete and newly imputed values together
imputed <- rbind(complete, merged)
sum(!complete.cases(imputed)) # should be 0 incomplete cases now
```

Now we can make another histogram and recalculate the mean and median:
```{r, fig.width=8, fig.height=4, echo=TRUE}
imp.steps.each.day <- tapply(imputed$steps, data$date, sum)

bwidth <- diff(range(imp.steps.each.day, na.rm=TRUE)) / 30
qplot(imp.steps.each.day, geom="histogram", binwidth=bwidth) +
    geom_rug(color="blue", alpha=0.7) +
    labs(x="Steps per day", y="Frequency",
         title="Histogram of total steps per day (post imputation)")

imp.mu <- mean(imp.steps.each.day, na.rm=TRUE)
imp.med <- median(imp.steps.each.day, na.rm=TRUE)
imp.mu
imp.med
```

So the mean is now **`r imp.mu`** and the median is **`r imp.med`**,
which are both different from the values we obtained before imputation
(`r mu`, `r med`, respectively).


## Are there differences in activity patterns between weekdays and weekends?

We'll add an boolean indicator variable *weekend* to our data.frame that
includes imputed values and make a plot to answer the question.

```{r, fig.width=8, fig.height=4, echo=TRUE}
wkend.days <- c("Saturday", "Sunday")
day.type <- factor(imputed$weekday %in% wkend.days)
imputed$day.type <- mapvalues(day.type, from=c("FALSE", "TRUE"), to=c("Weekday", "Weekend"))
imputed.means <- ddply(imputed, .(interval, day.type), summarise, mean=mean(steps, na.rm=TRUE))
head(imputed.means)
ggplot(imputed.means, aes(interval, mean)) + geom_line() + facet_grid(day.type ~ .) +
    labs(x="Interval", y="Steps", title="Activity by day type")
```

There appears to be a difference. For one thing, our subject appears to sleep
a bit later on the weekends (much like myself).
